\newpage
\TOCadd{Abstract}

\section*{\large Abstract}

%edit this text
\textcolor{red}{This needs to be edited still}
As the world moves away from fossil fuels there is a growing problem of inadequate shutdown procedures throughout the industry. Wells and silos become abandoned and leak oil into arable soil, ruining their quality for farmers and developers. Current soil remediation efforts are unable to accurately map and predict the spills and are unable to efficiently remediate the land. The advent of real-time soil sensing technology allows for the development of new prediction and mapping algorithms. The ensemble Kalman filter is a tool used in numerical weather prediction and could prove useful to soil modelling. These solutions require detailed physics-based models that described the movement of chemicals and organics in the soil. Additionally, implementation of data assimilation methods informed by the models is important. Below, the results of the ensemble Kalman filter on a diffusion and chemical reaction based soil model are analyzed. 

% This document is a possible Latex framework for a thesis or dissertation at UVic. It should work in the Windows, Mac and Unix environments. The content is based on the experience of one supervisor and graduate advisor. It explains the organization that can help write a thesis, especially in a scientific environment where the research contains experimental results as well. There is no claim that this is the \textit{best} or \textit{only} way to structure such a document. Yet in the majority of cases it serves extremely well as a sound basis which can be customized according to the requirements of the members of the supervisory committee and the topic of  research. Additionally some examples on using \LaTeX are included as a bonus for beginners.
